
\subsection*{Exercises for 5.1--5.4}
\begin{enumerate}
 \item

  Let \p{x,y} be variables and let the following keys explain the meanings of 
  some predicates and constants:
  \begin{lkey}
  \item[Nx] \p{x}is a narcissist

  \item[Lxy] \p{x} loves \p{y}

  \item[Px] is popular

  \item[d] Donald
   
  \item[f] Francis

  \item[m] Magdalena

  \end{lkey}

 Translate the following sentences into \lL[Q] (it is often helpful to translate 
 back into English to see if it comes out meaning the same as the original):

  \begin{enumerate}

   \item Donald is a narcissist.

   \item If Donald is a narcissist, then Donald loves Donald.

   \item Narcissists love themselves.

   \item If someone is a narcissist, they are not popular.

   \item Narcissists are not popular.

   \item Everyone loves Magdalena.

   \item Magdalena loves everyone.

   \item There is someone Francis loves.

   \item There is someone who loves Francis.

   \item Everyone loves someone.

   \item Someone is loved by everyone.

   \item If someone is loved by everyone, then everyone loves someone.

  \end{enumerate}
\answer{
 \begin{enumerate}
  \item \p{Nd}
  \item \p{Nd\limplies Ldd}
  \item \p{\lforall x(Nx\limplies Lxx)}
  \item \p{\lforall x(Nx\limplies\lnot Px)}
  \item \p{\lforall x(Nx\limplies\lnot Px)}
  \item \p{\lforall xLxm}
  \item \p{\lforall xLmx}
  \item \p{\lthereis xLfx}
  \item \p{\lthereis xLxf}
  \item \p{\lforall x\lthereis yLxy}
  \item \p{\lthereis x\lforall yLyx}
 \item \lbh{-p -f CXxUyLyxUxXyLxy}

 \end{enumerate}
}


 \item Consider the following formula:

  \lbh{-p -f XxKFxXyKFyKNGxyNXzKNGxkKNGyzFz}
 
 
	\begin{enumerate}

	 \item What is the scope of the left-most quantifier?

	 \item What is the scope of the right-most quantifier?

	 \item True of false: the formula contains free variables.

	\end{enumerate}
\answer{


 \begin{enumerate}

  \item The entire rest of the sentence.
  \item \lbh{-p -f KNGxkKNGyzFz}
 \item True. \p{k} is free (should have specified \p{k} is not a constant).
 \end{enumerate}

}

\item Consider the following formula (\p{x,y,z} are variables):

 \lbh{-p -f CXxFxUxCNFxXyKNGxzFy}


	\begin{enumerate}
	 \item Which variable in the formula is free?

	 \item What is the scope of the left-most existential quantifier?

	 \item What is the scope of the universal quantifier?

	\end{enumerate}

	\answer{
	 \begin{enumerate}
	\item \p{z}.
	\item Just the \p{Fx} left of the first \p{\limplies}.
	\item \lbh{-p -f CNFxXyKNGxzFy}
   \end{enumerate}
	}



  \end{enumerate}
